### 2020-02-11 manuscript.tex
Benchmark new set

\textbf{2. Time to convergence}. Two independent MCMC chains are run and the time is measured until: a) the absolute difference in clade posterior probability between the two chains is less than 0.05 for all clades, b) the Rubin-Gelman statistic $\hat{R}$ \cite{gelman1992inference} of every estimated parameter is less than 1.05, and c) the effective sample size \cite{rambaut2013tracer} of every estimated parameter is greater than 100 in each chain. \\

@@

For the latter two criteria, methodologies are benchmarked using the empirical datasets compiled and partitioned \cite{lanfear2016partitionfinder} by Lanfear \cite{lanfear2019Github}. Two datasets (the turtle data by Crawford et al. 2012 \cite{Crawford_2012}; and the songbird data by Moyle et al. 2016 \cite{Moyle_2016}) are used during the optimisation stage, where each setting is run $15 \times$ and the average statistics across the $15 \times 2 = 30$ chains are reported. All 25 datasets presented in \textbf{Table \ref{table:datasets}} are benchmarked together $1 \times$ each at the end of the protocol. Datasets and BEAST2 .xml templates are provided in the GitHub repository accompanying this article.

@@

 1  &  6  &  1/4/8/16  &  0.4/2.1/10.2/13.7  &  0.1/0.5/1/1.1  &  Richart 2015 \cite{Richart_2016} \\

@@

15  &  70  &  1/3  &  0.7/2.2  &  0.3/0.9  &  Kawahara 2013 \cite{Kawahara_2013} \\

@@

\textbf{19}  &  \textbf{106}  &  1/\textbf{4}/8/16  &  0.6/\textbf{3.1}/5.1/11.6  &  0.3/\textbf{1.2}/2/4.5  &  \textbf{Moyle 2016} \cite{Moyle_2016} \\

@@

 \hline
\end{tabular}
\caption{Empirical datasets used during benchmarking, sorted in increasing order of taxa count $N$. Number of partitions $P$, total alignment length $L$, and number of patterns $L_\text{eff}$ are also specified. Where a dataset is sampled more than once, the dimensions of its multiple samples are separated by `/'. The datasets in bold (Crawford \cite{Crawford_2012} and Moyle \cite{Moyle_2016}) are used during the optimisation protocol. }
\label{table:datasets}
\end{table}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/ESS.pdf}
\caption{\textbf{Rate parameterisation evaluation.} Comparison of ESS/hr (averaged across two independent MCMC chains) with respect to relevant terms -- $L$: likelihood, $p$: prior density, $r$: clock rate averaged across all leaves, $\sigma$: clock standard deviation. Each point is from one partition sample of the empirical data in \textbf{Table \ref{table:datasets}}.  }
\label{fig:parameterisationResults}
\end{figure}

@@

We selected the best operator variant by performing MCMC on 100 simulated datasets, where each MCMC employed all 96 NER/NERw variants. The acceptance rate of each operator is compared to that of the null operator $\text{NER}\{\}$ (ie. Narrow Exchange).

\textbf{Fig. \ref{fig:acceptanceRateScreening}} shows that NER variants which satisfy the genetic distances between nodes $B$ and $A$ (ie. $\mathcal{D}_{AB}$) or between $B$ and $C$ (ie. $\mathcal{D}_{BC}$) consistently perform worse than the standard Narrow Exchange operator, where $B$ is the node being interchanged from the $A$ branch to the $C$ branch (\textbf{Fig. \ref{fig:narrowexchange}}). This is an intuitive result. If the posterior distribution is relatively flat, and the data presents high uncertainty in the positioning of $B$, with respect to $A$ and $C$, then the topological rearrangement performed by Narrow Exchange will be favoured. However, this uncertainty in the \textit{topology} is likely coupled with uncertainty in the \textit{distance} between $B$ and $A$ or between $B$ and $C$. Thus, in this case, respecting the $\mathcal{D}_{AB}$ and $\mathcal{D}_{BC}$ constraints (by proposing branch rates) makes too many unnecessary changes to the state and the operator performs worse.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/acceptanceRates.pdf}
\caption{\textbf{Screening of NER and NERw variants by acceptance rate.} Left: comparison of NER variants with the null operator $\text{NER}\{\}$ (ie. Narrow Exchange). Each of the 48 operators are represented by a single point, uniquely encoded by the point stylings. The number of times each operator is proposed and accepted is compared with that of $\text{NER}\{\}$, and a Z-test is performed to determine whether there is a significant difference in the two acceptance rates.  This process is repeated for each of $100$ simulated datasets. The axes of each plot are the proportion of these $100$ simulations for which there is significant evidence that the operator is better than $\text{NER}\{\}$ (x-axis) or worse than $\text{NER}\{\}$ (y-axis). Right: comparison of NER with NERw. Each point is one NER/NERw variant from a single simulation. This plot shows that the random walk hinders the acceptance rates of all NER variants, including $\text{NER}\{\}$.    }
\label{fig:acceptanceRateScreening}
\end{figure}

\textbf{Fig. \ref{fig:acceptanceRateScreening}} also reveals a cluster of NER variants which -- under the conditions of the simulation --  performed better than the null operator $\text{NER}\{\}$ around 10\% of the time and performed worse around 40\% of the time. One such operator is  $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$. This variant conserves the genetic distance between all child nodes $A$, $B$, and $C$, and the grandparent node $E$. This is performed by proposing rates for $r_A$, $r_B$, and $r_C$ while obeying the distance constraints imposed by the operator. Exploring this operator further, we can see that $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ is at its best when there is a large variance in branch rate ie. when clock standard deviation $\sigma$ is high ($\sigma \gtrsim 0.3$), corresponding to data which is not clock-like. On the other hand, $\text{NER}\{\}$ is much preferred when the operator's acceptance rate is high ($\gtrsim 0.15$), which corresponds to poorer signal in the data.  Overall, $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ outperforms the standard Narrow Exchange operator provided that the data is not clock-like and contains strong signal.

@@

\subsubsection*{Advanced screening by congergence time}

>>>>

\textbf{2. Time to convergence}. Two independent MCMC chains are run and the time is measured until: a) the absolute difference in clade posterior probability between the two chains is less than 0.05 for all clades, b) the Rubin-Gelman statistic $\hat{R}$ \cite{gelman1992inference} of every estimated parameter is less than 1.05, and c) the effective sample size \cite{rambaut2018posterior} of every estimated parameter is greater than 100 in each chain. \\

@@

For the latter two criteria, methodologies are benchmarked using the empirical datasets compiled and partitioned \cite{lanfear2016partitionfinder} by Lanfear \cite{lanfear2019Github}. Four datasets (the spider data by Richart et al. 2015 \cite{Richart_2016}; the turtle data by Crawford et al. 2012 \cite{Crawford_2012}; the caterpillar dataset by Kawahara et al. 2013 \cite{Kawahara_2013}; and the songbird data by Moyle et al. 2016 \cite{Moyle_2016}) are used during the optimisation stage, where each setting is run $10 \times$ and the average statistics across the $10 \times 2 = 20$ chains are reported. All 25 datasets presented in \textbf{Table \ref{table:datasets}} are benchmarked together $1 \times$ each at the end of the protocol. Datasets and BEAST2 .xml templates are provided in the GitHub repository accompanying this article.

@@

\textbf{1}  &  \textbf{6}  &  1/4/8/\textbf{16}  &  0.4/2.1/10.2/\textbf{13.7}  &  0.1/0.5/1/\textbf{1.1}  &  \textbf{Richart 2015} \cite{Richart_2016} \\

@@

\textbf{15}  &  \textbf{70}  &  1/\textbf{3}  &  0.7/\textbf{2.2}  &  0.3/\textbf{0.9}  &  \textbf{Kawahara 2013} \cite{Kawahara_2013} \\

@@

\textbf{19}  &  \textbf{106}  &  \textbf{1}/4/8/16  &  \textbf{0.6}/3.1/5.1/11.6  &  \textbf{0.3}/1.2/2/4.5  &  \textbf{Moyle 2016} \cite{Moyle_2016} \\

@@

 \hline
\end{tabular}
\caption{Empirical datasets used during benchmarking, sorted in increasing order of taxa count $N$. Number of partitions $P$, total alignment length $L$, and number of patterns $L_\text{eff}$ are also specified. Where a dataset is sampled more than once, the dimensions of its multiple samples are separated by `/'. The datasets in bold (Richart 2015 \cite{Richart_2016}, Crawford \cite{Crawford_2012}, Kawahara \cite{Kawahara_2013}, and Moyle \cite{Moyle_2016}) are used during the optimisation protocol. }
\label{table:datasets}
\end{table}

@@

\begin{figure}[!h]
%\includegraphics[width=\textwidth]{Figures/ESS.pdf}
\caption{\textbf{Rate parameterisation evaluation.} Comparison of ESS/hr (averaged across two independent MCMC chains) with respect to relevant terms -- $L$: likelihood, $p$: prior density, $r$: clock rate averaged across all leaves, $\sigma$: clock standard deviation. Each point is from one partition sample of the empirical data in \textbf{Table \ref{table:datasets}}.  }
\label{fig:parameterisationResults}
\end{figure}

@@

We selected the best operator variant by performing MCMC on 300 simulated datasets, where each MCMC employed all 96 NER/NERw variants. Simulated datasets have $N=30$ taxa and an alignment with $L \sim \text{Uniform}(10^2, 10^4)$ sites. The acceptance rate of each operator is compared to that of the null operator $\text{NER}\{\}$ (ie. Narrow Exchange).

\textbf{Fig. \ref{fig:acceptanceRateScreening}} shows that NER variants which satisfy the genetic distances between nodes $B$ and $A$ (ie. $\mathcal{D}_{AB}$) or between $B$ and $C$ (ie. $\mathcal{D}_{BC}$) usually perform worse than the standard Narrow Exchange operator, where $B$ is the node being interchanged from the $A$ branch to the $C$ branch (\textbf{Fig. \ref{fig:narrowexchange}}). This is an intuitive result. If the posterior distribution is relatively flat, and the data presents high uncertainty in the positioning of $B$, with respect to $A$ and $C$, then the topological rearrangement performed by Narrow Exchange will be favoured. However, this uncertainty in the \textit{topology} is likely coupled with uncertainty in the \textit{distance} between $B$ and $A$ or between $B$ and $C$. Thus, in this case, respecting the $\mathcal{D}_{AB}$ and $\mathcal{D}_{BC}$ constraints (by proposing branch rates) makes too many unnecessary changes to the state and the operator performs worse.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/acceptanceRates.pdf}
\caption{\textbf{Screening of NER and NERw variants by acceptance rate.} Top left: comparison of NER variants with the null operator $\text{NER}\{\}$ (ie. Narrow Exchange). Each of the 48 operators are represented by a single point, uniquely encoded by the point stylings. The number of times each operator is proposed and accepted is compared with that of $\text{NER}\{\}$, and one-sided z-tests are performed to assess the statistical significance between the two acceptance rates ($p = 0.001$).  This process is repeated for each of $300$ simulated datasets. The axes of each plot are the proportion of these $300$ simulations for which there is evidence that the operator is better than $\text{NER}\{\}$ (x-axis) or worse than $\text{NER}\{\}$ (y-axis). Top right: comparison of NER and NERw acceptance rates. Each point is one NER/NERw variant from a single simulation. Bottom: relationship between the acceptance rates $\alpha$ of $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ and $\text{NER}\{ \}$ with the clock model standard deviation $\sigma$ and the number of patterns $L_\textit{eff}$.  Each point is a single simulation. }
\label{fig:acceptanceRateScreening}
\end{figure}

\textbf{Fig. \ref{fig:acceptanceRateScreening}} also reveals a cluster of NER variants which -- under the conditions of the simulation --  performed better than the null operator $\text{NER}\{\}$ around 25\% of the time and performed worse around 10\% of the time. One such operator is  $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$. This variant conserves the genetic distance between all child nodes $A$, $B$, and $C$, and the grandparent node $E$. This is performed by proposing rates for $r_A$, $r_B$, and $r_C$ while obeying the distance constraints imposed by the operator. Exploring this operator further, we can see that $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ is at its best when there is a large variance in branch rate ie. when clock standard deviation $\sigma$ is high ($\sigma \gtrsim 0.5$ for $N=30$), corresponding to data which is not clock-like. On the other hand, $\text{NER}\{\}$ is much preferred when the operator's acceptance rate is high ($\gtrsim 0.15$) -- corresponding with datasets with a small number of site patterns ($L_\textit{eff} \lesssim 500$ for $N=30$) and thus poor signal.  Overall, $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ outperforms the standard Narrow Exchange operator when the data is not clock-like and contains enough signal.

@@

\subsubsection*{Benchmarking convergence time}
